---
title: "Parenthood, Immigration and Further Education"
subtitle: "Data Cleaning/Prep"
author: "Nate Breznau, Johannes Bonnes and Sylvia Rahn"
output:
  word_document:
    toc: true
    toc_depth: '1'
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 1
  pdf_document:
    toc: true
    toc_depth: '1'
---

This code is setup up to also investigate Cycle 1, but we use only Cycle 2 here. 

## Setup   

```{r setup, echo = F, warning = F, message = F}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

packages <- c('tidyverse',
              'readr',
              'knitr',
              'here',
              'kableExtra',
              'countrycode',
              'readxl')

pacman::p_load(packages, character.only = T)

```

## Data

These data were combined from the PUFs of PIAAC using the routines in

https://github.com/nbreznau/ALE_degrees_institutions

```{r data}

#df <- read_csv(here("Data", "piaac_combined.csv")) 
df2 <- read_csv(here("Data", "piaac_combined_2.csv"))
```

## Recode

### Individual Dataset

 #### Cycle 1

```{r}

# df <- df %>%
#   # subset(PAIDWORK12 == 1 | !C_Q07 %in% c(4, 5, 7, 8, 97) | NFE12JR == 5)  %>%
#   mutate(female = case_when(GENDER_R == 2 ~ 1,
#                           GENDER_R == 1 ~ 0,
#                           TRUE ~ NA_real_),
#          age10 = case_when(AGEG10LFS == 1 ~ 20,
#                          AGEG10LFS == 2 ~ 30,
#                          AGEG10LFS == 3 ~ 40,
#                          AGEG10LFS == 4 ~ 50,
#                          AGEG10LFS == 5 ~ 60,
#                          TRUE ~ NA_real_
#                          ),
#          
#          FORBORN = ifelse(J_Q04a == "1", 0,
#                            ifelse(J_Q04a == "2", 1, NA)),
#          born_in_country_mother = ifelse(J_Q06a == "1", 1,
#                                   ifelse(J_Q06a == "2", 0, NA)), # includes guardian if not with biological parents (2nd wave does not)
#          born_in_country_father = ifelse(J_Q07a == "1", 1,
#                                   ifelse(J_Q07a == "2", 0, NA)),
#          FORLANG = case_when(as.numeric(NATIVESPEAKER) == 1 ~ 0,
#                              as.numeric(NATIVESPEAKER) == 0 ~ 1,
#                              .default = NA_real_),
#          BORNLANG = case_when(NATIVESPEAKER == 2 & born_in_country == 1 ~ 1,
#                               NATIVESPEAKER == 1 & born_in_country == 1 ~ 2,
#                               NATIVESPEAKER == 2 & born_in_country == 0 ~ 3,
#                               NATIVESPEAKER == 1 & born_in_country == 0 ~ 4,
#                               .default = NA_real_),
#          # fix mising EDLEVEL3 in Austria
#          EDCAT6 = as.numeric(EDCAT6),
#          et = case_when(EDCAT6 == 2 | EDCAT6 == 3 ~ 2,
#                         EDCAT6 == 4 | EDCAT6 == 5 | EDCAT6 == 6 | EDCAT6 == 7 ~ 3,
#                         CNTRYID != 40 ~ NA,
#                         .default = EDCAT6),
#          ed3 = ifelse(CNTRYID == 40, et, EDLEVEL3),
#          ed_currently = B_Q02a, 
#          children = case_when(J_Q03a == 2 ~ 0,
#                               J_Q03a == 1 ~ 1,
#                               TRUE ~ 0),
#          #building an equivalent 'num of CVET courses' measure
#          num1 = as.numeric(B_Q12b), #open or distance education?
#          num2 = as.numeric(B_Q12d), #on-the-job training
#          num3 = as.numeric(B_Q12f), #seminars and workshops
#          num4 = as.numeric(B_Q12h), #courses other not reported
#          NFE_number = num1 + num2 + num3 + num4,
#     skill_reading = rowMeans(select(., starts_with("PVLIT")), na.rm = TRUE),
#     skill_numeracy = rowMeans(select(., starts_with("PVNUM")), na.rm = TRUE),
#         group_07 = case_when(children == 1 & female == 1 & FORBORN == 0 ~ 4,
#                          children == 0 & female == 1 & FORBORN == 0 ~ 3,
#                          children == 1 & female == 1 & FORBORN == 1 ~ 2,
#                          children == 0 & female == 1 & FORBORN == 1 ~ 1,
#                          .default = NA_real_),
#     literacy_level = case_when( # PIAAC's own 5-category system
#       skill_reading < 176 ~ 0,
#       skill_reading < 226 ~ 1,
#       skill_reading < 276 ~ 2,
#       skill_reading < 326 ~ 3,
#       skill_reading > 325 ~ 4,
#       TRUE ~ NA_real_
#     ),
#     numeracy_level = case_when(
#       skill_numeracy < 176 ~ 0,
#       skill_numeracy < 226 ~ 1,
#       skill_numeracy < 276 ~ 2,
#       skill_numeracy < 326 ~ 3,
#       skill_numeracy > 325 ~ 4,
#       TRUE ~ NA_real_
#     )
#   ) %>%
#   select(iso3c, skill_reading, skill_numeracy, literacy_level, numeracy_level, NFE12, NFE12JR, female, children, FORLANG, FORBORN, BORNLANG, group_07, NFE_number, ed3, ed_currently, SPFWT0) 
# 
# df <- df %>%
#   group_by(iso3c) %>%
# # create a weighting variable that is accurate at the country-level, that does not compare populations across countries
#   mutate(aweight = SPFWT0 / min(SPFWT0, na.rm = T)) %>%
#   ungroup()
#   

```

#### Cycle 2

B2_Q05a - **Are you currently studying for any kind of formal qualification?**
c(Yes = 1, No = 2) 

B2_Q22 - **In the last 12 months, were there any training activities you wanted to participate in but did not?**
c(Yes = 1, No = 2) 

B2_Q23 - **Which of the following reasons prevented you from participating in these training activities? Please indicate the most important reason.** 
c('I did not find any suitable training activity' = 1, 
'I had no time due to family responsibilities' = 2, 
'I had no time due to work-related reasons' = 3, 
'I did not have the prerequisites' = 4, 
'Lack of employer’s support' = 5, 
'Training activity would have been too expensive' = 6, 
'Training activity took place at an inconvenient time or location' = 7, 
'Training activity was canceled or postponed' = 8, 
'Something unexpected came up that prevented me from participating' = 9, 
'Other reason' = 10
) 

H2_Q19a - **Overall, which of the following statements best describes your skills in relation to what is required to do your job?**
c('My skills are higher than required by my job' = 1, 
'My skills are matched to what is required by my job' = 2, 
'Some of my skills are lower than what is required by my job and need to be further developed' = 3)

```{r data}
df2 <- df2 %>%
  mutate(
    iso3c = countrycode(as.numeric(CNTRYID), "iso3n", "iso3c"),
         iso3c = ifelse(as.numeric(CNTRYID_E) == 7561, "CHE",
                 ifelse(as.numeric(CNTRYID_E) == 7562, "CHE",
                 ifelse(as.numeric(CNTRYID_E) == 7563, "CHE", 
                 ifelse(as.numeric(CNTRYID_E) == 152, "CHL",
                 ifelse(as.numeric(CNTRYID_E) == 203, "CZE", iso3c))))),
    skill_reading = rowMeans(select(., starts_with("PVLIT")), na.rm = TRUE),
    skill_numeracy = rowMeans(select(., starts_with("PVNUM")), na.rm = TRUE),
    literacy_level = case_when( # PIAAC's own 5-category system
      skill_reading < 176 ~ 0,
      skill_reading < 226 ~ 1,
      skill_reading < 276 ~ 2,
      skill_reading < 326 ~ 3,
      skill_reading > 325 ~ 4,
      TRUE ~ NA_real_
    ),
    numeracy_level = case_when(
      skill_numeracy < 176 ~ 0,
      skill_numeracy < 226 ~ 1,
      skill_numeracy < 276 ~ 2,
      skill_numeracy < 326 ~ 3,
      skill_numeracy > 325 ~ 4,
      TRUE ~ NA_real_
    ),
    FORBORN = case_when(BORNLANG == 1 | BORNLANG == 2 ~ 0,
                        BORNLANG == 3 | BORNLANG == 4 ~ 1,
                        .default = NA),
    FORLANG = case_when(BORNLANG == 1 | BORNLANG == 3 ~ 0,
                        BORNLANG == 2 | BORNLANG == 4 ~ 1,
                        .default = NA),
    EDCAT6 = as.numeric(EDCAT6_TC1),
    EDCAT7 = as.numeric(EDCAT7_TC1),
    ed3 = case_when(EDCAT6 == 2 | EDCAT6 == 3 ~ 2,
                    EDCAT6 == 4 | EDCAT6 == 5 | EDCAT6 == 6 | EDCAT6 == 7 ~ 3, 
                    .default = EDCAT6),
    ed3_backup = case_when(EDCAT7 == 1 ~ 1,
                            EDCAT7 > 1 & EDCAT7 < 5 ~ 2,
                            EDCAT7 > 4 ~ 3,
                            TRUE ~ NA_real_),
    ed3 = ifelse(is.na(ed3), ed3_backup, ed3),
    ed_currently = ifelse(B2_Q05a == "1", 1, 0),
    ed_currently = ifelse(is.na(ed_currently), 0, ed_currently), # recode missing to 0 for parsimony
    NFE_barrier = ifelse(B2_Q22 == "1", 1, 0),
    NFE_barrier = ifelse(is.na(NFE_barrier), 0, NFE_barrier), # recode missing to 0 for parsimony
    underskilled = case_when(H2_Q19a == "3" ~ 1,
                             .default = 0),
    year = 2022,
    A2_N02_T = as.numeric(A2_N02_T),
    female = ifelse(A2_N02_T == 2, 1, 0),
    female_2 = ifelse(GENDER_R == 2, 1, 0),
    female = ifelse(is.na(female), female_2, female),
    children = ifelse(J2_Q03a == 1, 1, 0),
    children = ifelse(is.na(children), 0, children),
    group_07 = case_when(children == 1 & female == 1 & FORBORN == 0 ~ 4,
                         children == 0 & female == 1 & FORBORN == 0 ~ 3,
                         children == 1 & female == 1 & FORBORN == 1 ~ 2,
                         children == 0 & female == 1 & FORBORN == 1 ~ 1,
                         .default = NA_real_),
    NFE_number = as.numeric(NFE12NR),
    NFE12JR = NFE12JRC2,
    NFE12 = ifelse(B2_Q08a == "2", "0",
            ifelse(B2_Q08a == "1", "1", NA)), # participated in any training last 12 months
    NFE12 = ifelse(is.na(NFE12), NFE12C2, NFE12),
    NFE12 = as.numeric(NFE12),
    NFE12JRa = as.numeric(NFE12JRC2), # note that this variable is coded differently than in wave one. This wave, those who did not participate in any training are coded as missing, so that this variable is only the percentage of those who reported training that reported it was job-related. This has to be fixed to compare them
    NFE12JR = ifelse(!is.na(NFE12) & is.na(NFE12JRa), "0", as.character(NFE12JRa))) %>%
  select(iso3c, skill_reading, skill_numeracy, literacy_level, numeracy_level,NFE12, NFE12JR, NFE_barrier, underskilled, female, children, FORLANG, FORBORN, BORNLANG, group_07, NFE_number, ed3, ed_currently, SPFWT0) 

df2 <- df2 %>%
  group_by(iso3c) %>%
# create a weighting variable that is accurate at the country-level, that does not compare populations across countries
  mutate(aweight = SPFWT0 / min(SPFWT0, na.rm = T)) %>%
  ungroup()

```

### Country Indicators

#### GDP

https://dataverse.nl/dataset.xhtml?persistentId=doi:10.34894/INZBF2

```{r gdp}
#get all countries
clist <- unique(df2$iso3c)


df_gdp <- read_excel(here("Data", "mpd2023_web.xlsx"), sheet = "Full data", col_types = c("text", "text", "text", "numeric", "numeric", "numeric")) %>%
  mutate(iso3c = countrycode) %>%
  subset(iso3c %in% clist) %>%
  subset(year > 2018) %>%
  select(iso3c, gdppc, pop) %>%
  group_by(iso3c) %>%
  summarise_all(mean, na.rm = T) %>%
  ungroup()

```



## Save

```{r }

df2 <- df2 %>%
  left_join(df_gdp, by = c("iso3c"))
  
#write_rds(df, here("Data", "df.RDS"))

write_rds(df2, here("Data", "df2.RDS"))
```

## Citations

```{r}

packages %>%
  map(citation) %>%
  print(style = "text")

```


## Colophon

```{r}

sessionInfo()

```


