---
title: "06 Simulated interventions (Table 1)"
subtitle: "The 'Immigration Issue' Misnomer in Human Capital Research: Intersectional Inequalities in Adult Education and Literacy"
author: "Author omitted for blind peer review"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
always_allow_html: true
---

**What this script does.** It reproduces the simulator app's three
intervention scenarios for three fixed pairs of target groups (X and Y)
and reports the results as Table 1 of the main text.

| Output file (in `Results/`) | Paper |
|---|---|
| `Table_1.docx/.csv` | Table 1 (simulated gains, 3 group pairs x 3 scenarios) |
| `Table_1_data.csv` | Country values behind the table |

**Input.** `Shiny/precomp_spec_country.rds` from 04 (the same file the app
reads), so the table and the app always agree.

**The scenarios** (as in the app, `Shiny/app.R`). Each treats a share of
the group's members who had no ALE in the last 12 months:

- *Strict parity*: 50% of the untreated in X; in Y the same number of
  people (as a share of the population), or all of Y's untreated if Y has
  fewer.
- *Limited reach*: 25% of the untreated in X; in Y up to
  `limited_multiplier` times that number of people (set below; the app
  uses 1.5).
- *Blank check*: 75% of the untreated in each group, no limit.

The gain is the rise in the country's average literacy, in points, when
the treated get the group's effect of ALE (M5) plus the country's own
deviation in the ALE slope ("context effect", as with the app's default
setting), floored at zero.

## Setup

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(here)
source(here("R", "load_packages.R"))

packages <- c("tidyverse", "here", "knitr", "flextable")
load_packages(packages)

precomp <- read_rds(here("Shiny", "precomp_spec_country.rds"))

# TRUE = include each country's context effect (the app's default)
use_context <- TRUE

# Limited reach: Y may treat this many times as many people as X. The app
# uses 1.5; 2 splits the countries of pair A about evenly (see the
# break-even table at the end).
limited_multiplier <- 2
```

## Group pairs

A group is given as in the app: 1 = yes, 0 = no, `NA` = both.

- A: foreign-born non-native-speaking mothers vs native-born
  native-speaking mothers (both education levels).
- B: the same comparison for men without children.
- C: the group with the largest effect of ALE in M5 vs the group with the
  smallest, among the fully specified groups of Figure 3 (native-born
  non-native speakers are left out there, as too few per country).

```{r groups}
make_spec_id <- function(female = NA, FORBORN = NA, FORLANG = NA,
                         children = NA, ed3_tertiary = NA) {
  v <- function(x) ifelse(is.na(x), "all", x)
  sprintf("f=%s;b=%s;l=%s;c=%s;e=%s", v(female), v(FORBORN), v(FORLANG),
          v(children), v(ed3_tertiary))
}

# describe a spec_id in words
spec_label <- function(id) {
  x <- setNames(sub(".*=", "", strsplit(id, ";")[[1]]), c("f", "b", "l", "c", "e"))
  pick <- function(v, yes, no) switch(v, "1" = yes, "0" = no, NULL)
  parent <- if (x["c"] == "all") NULL else
    if (x["f"] == "1") pick(x["c"], "mother", "woman without children") else
    if (x["f"] == "0") pick(x["c"], "father", "man without children") else
    pick(x["c"], "parent", "no children")
  paste(c(pick(x["b"], "Foreign-born", "Native-born"),
          pick(x["l"], "non-native speaker", "native speaker"),
          if (is.null(parent)) pick(x["f"], "female", "male") else parent,
          pick(x["e"], "tertiary degree", "no tertiary degree")),
        collapse = ", ")
}

# effect of ALE (M5) for every fully specified group in Figure 3
full_groups <- precomp %>%
  filter(!grepl("all", spec_id), !grepl("b=0;l=1", spec_id)) %>%
  group_by(spec_id) %>%
  summarise(AME = first(AME), SE = first(SE),
            pct_population = 100 * mean(pct_group), .groups = "drop") %>%
  arrange(desc(AME))
kable(full_groups, digits = 2)

pairs <- tribble(
  ~group, ~x_id, ~y_id,
  "A", make_spec_id(female = 1, FORBORN = 1, FORLANG = 1, children = 1),
       make_spec_id(female = 1, FORBORN = 0, FORLANG = 0, children = 1),
  "B", make_spec_id(female = 0, FORBORN = 1, FORLANG = 1, children = 0),
       make_spec_id(female = 0, FORBORN = 0, FORLANG = 0, children = 0),
  "C", first(full_groups$spec_id), last(full_groups$spec_id)) %>%
  mutate(x_label = map_chr(x_id, spec_label),
         y_label = map_chr(y_id, spec_label))

kable(pairs)
```

## Simulation

```{r simulate}
scenarios <- tribble(
  ~scenario,        ~frac_x, ~y_rule,
  "Strict parity",  0.50,    "match",
  "Limited reach",  0.25,    "match_k",
  "Blank check",    0.75,    "same_share")

group_values <- function(id, suffix) {
  precomp %>%
    filter(spec_id == id) %>%
    transmute(iso3c,
              ATE = AME,
              gain_all = if (use_context) skill_gain_ctx else skill_gain_base,
              pct_group, pct_noale = pct_group_noale) %>%
    rename_with(~ paste0(.x, suffix), -iso3c)
}

simulate_pair <- function(x_id, y_id) {
  left_join(group_values(x_id, "_X"), group_values(y_id, "_Y"), by = "iso3c") %>%
    crossing(scenarios) %>%
    mutate(
      treated_share_X = pct_noale_X * frac_x,
      frac_y = case_when(
        y_rule == "same_share" ~ frac_x,
        pct_noale_Y <= 0       ~ 0,
        y_rule == "match"      ~ pmin(1, treated_share_X / pct_noale_Y),
        y_rule == "match_k"    ~ pmin(1, limited_multiplier * treated_share_X / pct_noale_Y)),
      treated_share_Y = pct_noale_Y * frac_y,
      # gain_all_* is the gain if every untreated member is treated, so a
      # share of them gives that share of the gain
      gain_X = gain_all_X * frac_x,
      gain_Y = gain_all_Y * frac_y)
}

sims <- pairs %>%
  mutate(res = map2(x_id, y_id, simulate_pair)) %>%
  unnest(res) %>%
  mutate(scenario = factor(scenario, levels = scenarios$scenario)) %>%
  arrange(group, scenario, iso3c)

write_csv(sims, here("Results", "Table_1_data.csv"))
```

## Table 1

The gains are changes in the average literacy of each country's whole
adult population (not only of the treated group), averaged over countries
with each country counting equally. The difference shows whether
targeting X instead of Y raises population literacy more; the last column
counts the countries where it does.

```{r table1}
table_1 <- sims %>%
  group_by(group, x_label, y_label, scenario) %>%
  summarise(ATE_X = first(ATE_X), ATE_Y = first(ATE_Y),
            # count before gain_X and gain_Y are replaced by their means
            countries_X_more = sprintf("%d of %d", sum(gain_X > gain_Y), n()),
            treated_pct_X = 100 * mean(treated_share_X),
            treated_pct_Y = 100 * mean(treated_share_Y),
            gain_X = mean(gain_X), gain_Y = mean(gain_Y),
            .groups = "drop") %>%
  mutate(gain_diff = gain_X - gain_Y) %>%
  mutate(across(c(ATE_X, ATE_Y), ~ round(.x, 1)),
         across(c(treated_pct_X, treated_pct_Y), ~ round(.x, 2)),
         across(c(gain_X, gain_Y, gain_diff), ~ round(.x, 2)))

table_1_out <- table_1 %>%
  transmute(Group = paste0(group, ". X: ", x_label, " / Y: ", y_label),
            Scenario = scenario,
            `% of population treated, X` = treated_pct_X,
            `% of population treated, Y` = treated_pct_Y,
            `Change in population literacy, treating X` = gain_X,
            `Change in population literacy, treating Y` = gain_Y,
            `Difference (X minus Y)` = gain_diff,
            `Countries where X gains more` = countries_X_more)

write_csv(table_1_out, here("Results", "Table_1.csv"))
flextable(table_1_out) %>%
  merge_v(j = "Group") %>%
  valign(j = "Group", valign = "top") %>%
  fontsize(size = 8, part = "all") %>%
  width(j = "Group", width = 2.2) %>%
  width(j = 2:8, width = 0.8) %>%
  hline(i = cumsum(rle(table_1_out$Group)$lengths)) %>%
  set_caption(paste0("Table 1. Simulated change in average population literacy (points) from ALE interventions targeting group X or group Y (limited reach: Y up to ", limited_multiplier, " times as many people as X)")) %>%
  save_as_docx(path = here("Results", "Table_1.docx"))
kable(table_1_out)

# effects of ALE for each group (M5), for the table note
table_1 %>% distinct(group, x_label, ATE_X, y_label, ATE_Y) %>% kable()
```

### Break-even multiplier

How many times as many people in Y as in X must be treated for Y to raise
population literacy as much as X (ignoring the limit of Y's untreated)?
Per country this is the gain per treated person in X divided by that in
Y; the table gives the median and range over countries. A limited-reach
multiplier near the median splits the countries about evenly.

```{r breakeven}
breakeven <- sims %>%
  filter(scenario == "Strict parity", pct_noale_X > 0, pct_noale_Y > 0,
         gain_all_Y > 0) %>%
  mutate(multiplier = (gain_all_X / pct_noale_X) / (gain_all_Y / pct_noale_Y)) %>%
  group_by(group) %>%
  summarise(countries = n(), median = median(multiplier),
            min = min(multiplier), max = max(multiplier), .groups = "drop")

write_csv(breakeven, here("Results", "Table_1_breakeven.csv"))
kable(breakeven, digits = 2)
```

## Package citations

```{r citations}
packages %>% map(citation) %>% print(style = "text")
```

## Session information

```{r session}
sessionInfo()
```
