---
title: "01 Data Setup"
subtitle: "The 'Immigration Issue' Misnomer in Human Capital Research: Intersectional Inequalities in Adult Education and Literacy"
author: "Author omitted for blind peer review"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

**What this script does.** It turns the combined PIAAC Cycle 2 file into the
analysis file used by every later script. It creates the analysis variables
(literacy, adult learning, sex, birth country, language, parenthood,
education), marks the analysis sample, adds country GDP, and saves
`Data/df2.RDS`.

**Inputs.**

- `Data/piaac_cycle2.rds`, made by `00_Import.R` from the OECD public-use
  files (PIAAC Cycle 2, data collected 2022–23).
- `Data/mpd2023_web.xlsx`, the Maddison Project Database 2023
  (<https://dataverse.nl/dataset.xhtml?persistentId=doi:10.34894/INZBF2>).

**Output.** `Data/df2.RDS`: one row per respondent in all 30 countries,
with `in_sample = 1` for the cases used in the analysis.

Only Cycle 2 is used. PIAAC Cycle 1 (2012–17) asked about adult learning
differently and is not part of the analysis.

## Setup

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(here)
source(here("R", "load_packages.R"))

packages <- c("tidyverse", "here", "readxl", "knitr")
load_packages(packages)
```

## Load the combined PIAAC file

```{r load}
raw <- read_rds(here("Data", "piaac_cycle2.rds"))

cat(nrow(raw), "respondents in", n_distinct(raw$iso3c), "countries\n")
```

## Recode

Each recode below keeps PIAAC's missing values as `NA` instead of turning
them into a substantive value, so that every case dropped later is visible
in the sample-flow table.

**Sex.** `GENDER_R` (1 male, 2 female). It agrees with the questionnaire item
`A2_N02_T` wherever both exist and is available for more respondents.

**Foreign-born and non-native speaker.** `BORNLANG` combines birth country
and first language: 1 native-born and native language, 2 native-born and
foreign language, 3 foreign-born and native language, 4 foreign-born and
foreign language. Denmark does not release `BORNLANG` in its public-use
file, so foreign-born status there falls back on `A2_Q03a` (born in this
country, 1 yes / 2 no); language stays missing for Denmark.

**Parenthood.** `J2_Q03a` (has children, 1 yes / 2 no).

**Education.** `EDCAT6_TC1`, PIAAC's six-category attainment variable:
1 lower secondary or less, 2 upper secondary, 3 post-secondary
non-tertiary, 4 tertiary professional degree, 5 bachelor, 6 master or
research. `ed3` groups these into primary/lower secondary (1), upper and
post-secondary (2) and tertiary (3). The models use only the tertiary
indicator. (`EDCAT7_TC1` gives the same tertiary split but is not released
for Austria, Switzerland, Norway and Sweden.)

**Currently in formal education.** `B2_Q05a` (1 yes / 2 no). Respondents
still studying are excluded from the analysis.

**Adult learning and education (ALE).** `NFE12C2`, PIAAC's derived
indicator of participation in non-formal education in the last 12 months
(0/1). It is identical to the questionnaire item `B2_Q08a` wherever both
exist and covers more respondents. `NFE12NR` is the number of activities.

**Literacy.** PIAAC reports literacy as ten plausible values
(`PVLIT1`–`PVLIT10`). `skill_reading` is their average, used as the
outcome in the regression models; descriptive statistics in 02 use the ten
plausible values separately.

```{r recode}
df2 <- raw %>%
  mutate(
    # literacy and numeracy: mean of the plausible values (see note above)
    skill_reading  = rowMeans(across(PVLIT1:PVLIT10)),
    skill_numeracy = rowMeans(across(PVNUM1:PVNUM10)),
    # PIAAC proficiency levels: below 1 (0) to level 4 and above (4)
    literacy_level = cut(skill_reading, c(-Inf, 176, 226, 276, 326, Inf),
                         labels = FALSE, right = FALSE) - 1,

    # sex
    female = case_when(GENDER_R == 2 ~ 1, GENDER_R == 1 ~ 0),
    male   = 1 - female,

    # birth country and language
    FORBORN = case_when(BORNLANG %in% 1:2 ~ 0,
                        BORNLANG %in% 3:4 ~ 1,
                        is.na(BORNLANG) & A2_Q03a == 1 ~ 0,
                        is.na(BORNLANG) & A2_Q03a == 2 ~ 1),
    FORBORN_not = 1 - FORBORN,
    FORLANG = case_when(BORNLANG %in% c(1, 3) ~ 0,
                        BORNLANG %in% c(2, 4) ~ 1),

    # parenthood
    children = case_when(J2_Q03a == 1 ~ 1, J2_Q03a == 2 ~ 0),

    # education
    ed3 = case_when(EDCAT6_TC1 == 1 ~ 1,
                    EDCAT6_TC1 %in% 2:3 ~ 2,
                    EDCAT6_TC1 %in% 4:6 ~ 3),
    tertiary = as.numeric(ed3 == 3),
    ed_currently = case_when(B2_Q05a == 1 ~ 1, B2_Q05a == 2 ~ 0),

    # adult learning in the last 12 months
    NFE12      = NFE12C2,
    NFE_number = NFE12NR,
    # job-related ALE; non-participants (missing in Cycle 2) are coded 0
    NFE12JR    = if_else(NFE12 == 0, 0, NFE12JRC2),

    # other items kept for robustness checks
    NFE_barrier  = case_when(B2_Q22 == 1 ~ 1, B2_Q22 == 2 ~ 0),
    underskilled = case_when(H2_Q19a == 3 ~ 1, H2_Q19a %in% 1:2 ~ 0),
    age10 = AGEG10LFS
  )
```

### Intersectional groups

The 16 groups cross sex, birth country, language and parenthood. `group16`
names each group the same way the figures do, e.g.
`female_fb_notnatlang_child` is a foreign-born mother who is not a native
speaker. It is missing when any of the four characteristics is missing.

```{r groups}
df2 <- df2 %>%
  mutate(group16 = if_else(
    is.na(female) | is.na(FORBORN) | is.na(FORLANG) | is.na(children),
    NA_character_,
    paste(if_else(female == 1, "female", "male"),
          if_else(FORBORN == 1, "fb", "not"),
          if_else(FORLANG == 1, "notnatlang", "natlang"),
          if_else(children == 1, "child", "nochild"),
          sep = "_")))

count(df2, group16) %>% kable()
```

### Weight for the multilevel models

`aweight` rescales each respondent's final weight by the smallest weight in
their country. It is kept because `03_Analyses.Rmd` uses it for the sample
bias check. The descriptive statistics in 02 use the final weight `SPFWT0`
and its replicates instead.

```{r aweight}
df2 <- df2 %>%
  group_by(iso3c) %>%
  mutate(aweight = SPFWT0 / min(SPFWT0)) %>%
  ungroup()
```

## Analysis sample

The analysis uses adults who are not in formal education and have valid
values on every model variable. The table shows how many cases each step
removes.

```{r sample}
model_vars <- c("skill_reading", "NFE12", "female", "FORBORN", "FORLANG",
                "children", "ed3")

df2 <- df2 %>%
  mutate(in_sample = as.numeric(
    ed_currently %in% 0 & if_all(all_of(model_vars), ~ !is.na(.x))))

sample_flow <- tibble(
  step = c("All respondents",
           "Not currently in formal education",
           "Valid on all model variables (analysis sample)"),
  n = c(nrow(df2),
        sum(df2$ed_currently %in% 0),
        sum(df2$in_sample)),
  countries = c(n_distinct(df2$iso3c),
                n_distinct(df2$iso3c[df2$ed_currently %in% 0]),
                n_distinct(df2$iso3c[df2$in_sample == 1])))

kable(sample_flow)
```

Missing values by country (percent of respondents). Denmark's public-use
file does not include language, education or current studies, so Denmark
drops out of the analysis sample entirely.

```{r missing_by_country}
df2 %>%
  group_by(iso3c) %>%
  summarise(n = n(),
            across(c(all_of(model_vars), ed_currently),
                   ~ round(100 * mean(is.na(.x)))),
            in_sample = sum(in_sample)) %>%
  kable()
```

## Country indicators: GDP per capita

GDP per capita (2011 US$, PPP) and population are averaged over
2019–2022, the last four years in the Maddison database. GDP is capped at
60,000 so that the three outliers (Switzerland, Norway and Singapore) do
not dominate the between-country slope, then divided by 1,000 and logged.

```{r gdp}
df_gdp <- read_excel(here("Data", "mpd2023_web.xlsx"), sheet = "Full data") %>%
  filter(countrycode %in% unique(df2$iso3c), year >= 2019) %>%
  group_by(iso3c = countrycode) %>%
  summarise(gdppc = mean(gdppc, na.rm = TRUE),
            pop   = mean(pop, na.rm = TRUE),   # thousands
            .groups = "drop")

df2 <- df2 %>%
  left_join(df_gdp, by = "iso3c") %>%
  mutate(gdppc_cap   = pmin(gdppc, 60000),
         gdppc_k     = gdppc_cap / 1000,
         gdppc_k_log = log(gdppc_k))

df_gdp %>% mutate(gdppc = round(gdppc)) %>% kable()
```

## Save

Only the variables used later are kept, along with the plausible values and
the full set of replicate weights.

```{r save}
df2 <- df2 %>%
  select(iso3c, CNTRYID, SEQID, in_sample,
         skill_reading, skill_numeracy, literacy_level,
         PVLIT1:PVLIT10, PVNUM1:PVNUM10,
         NFE12, NFE12JR, NFE_number, NFE_barrier, underskilled,
         female, male, FORBORN, FORBORN_not, FORLANG, BORNLANG, children,
         group16, ed3, tertiary, ed_currently, age10,
         gdppc, pop, gdppc_cap, gdppc_k, gdppc_k_log,
         aweight, SPFWT0, SPFWT1:SPFWT80, VEMETHOD, VEFAYFAC, VENREPS)

write_rds(df2, here("Data", "df2.RDS"))
cat("Saved Data/df2.RDS:", nrow(df2), "rows,", ncol(df2), "columns\n")
```

## Package citations

```{r citations}
packages %>% map(citation) %>% print(style = "text")
```

## Session information

```{r session}
sessionInfo()
```
